Towards open collaboration in biomedical and health research: An open collaborative scoping review

### 2025-01-08 doc/paper.qmd
docs: :pencil2: small formatting fixes for website

---
title: "Towards open collaboration in biomedical and health research: An open collaborative scoping review"
date: last-modified
author: 
  - name: Hannah Chatwin
    affiliations: 
      - ref: au
  - name: Mario Garcia
    affiliations: 
      - ref: ku
  - name: Daniel B. Ibsen
    affiliations: 
      - ref: sdca
      - ref: au
  - name: Luke W. Johnston
    affiliations: 
      - ref: sdca
      - ref: au
affiliations:
  - id: sdca
    name: Steno Diabetes Center Aarhus
    country: Denmark
  - id: au
    name: Aarhus University
    country: Denmark
  - id: ku
    name: Copenhagen University
    country: Denmark
format: html
bibliography: references.bib
csl: vancouver.csl
---

@@

The full protocol for this scoping review was uploaded to the Open
Science Framework [@Johnston2022].

As with any research project, things evolve compared to what was
originally intended or planned. In this section we describe what was
changed from the original, uploaded protocol.

### Document types

We originally intended on searching websites, blogs, and other online
resources, but it was very difficult to programmatically and
systemtically achieve this. So we ended up only searching for scholarly
work that was indexed in databases.

### Source databases

We originally aimed to review individual archives that included PubMed,
bioRxiv, Scopus, and several others (the full list is in the protocol),
and to use R packages with web API connections to each of these sources
to programmatically extract the sources we wanted. However we
encountered substantial issues that completely changed how we actually
found and extracted the sources.

The first challenge we encountered was that, while most of the source
databases had (hypothetically) exposed APIs along with R packages
available to access them, they didn't always work well or had
complicated instructions for actually using them. For instance, the
preprint repositories bioRxiv and medRxiv didn't have an up to date R
package to access the preprints nor was the web API well described, nor
well designed. In order to effectively use it, you need to download the
entire database of preprints locally before being able to search for the
preprints you want. This made it effectively impossible to use.

The second challenge we encountered was a difference in results between
using the web API compared to using the web search interface for some
source databases. There were some small differences in search results
between the API query compared to the web-based query. We couldn't
identify or understand how these differences happened or way, though
they were quite small differences (less than 10 out of thousands of
scholarly works from the search results).

The third challenge we had was that some source databases, like the Web
of Science or Scopus, didn't work consistently between the authors. In
order to use the web API with the R packages, you need to generate a
token for authentication. However, for some authors, the token worked
fine, and for others, it didn't work at all, with no explanation or
error message. Given our interest in and aim for reproducibility, since
the same code ran differently between authors, we decided to exclude
these source databases.

Given these challenges however, we unintentionally found
[OpenAlex](https://openalex.org/), which is a public database of
scholarly output that aggregates dozens of other source databases into
one, easy interface. This resource also has an R package called
`{openalexR}` [@Aria2024] with a very well designed interface, which
simplified or made redundant much of our code we wrote. It also made it
unnecessary to use the individual R packages for each source database
that we originally listed in the protocol. By using this archive, it
resolved all our challenges and barriers we had before. For a list of
where they get their source data from, see their ["About the
data"](https://help.openalex.org/hc/en-us/articles/24397285563671-About-the-data)
page.

### Inclusion and exlucion criteria

While we didn't change our search terms for the initial search and
extraction, we found that our search terms were not precise enough and
got many irrelevant scholarly works. The full R code to filter down the
search results is kept in the
[`R/exlusions.R`](https://github.com/science-collective/scoping-review/blob/main/R/exclusions.R)
file on the GitHub repository. For example, we extracted a large number
of scholarly works describing surgery such as "open wound", electronics
such as "open circuit", or environmental such as "open water" that we
had to post-process and exclude.

### Search period

We were not able to dedicate as much time to this project as we had
initially planned, so our "stopping rule" for end date didn't match with
our actual search period. The actual end date for when we did the last
search results extraction was
[2024-02-29](https://github.com/science-collective/scoping-review/commit/f288d3f23d028e77ad1081e0cfe3583ce4dcb8b6).

@@

### Scoping review identifies only 11 papers discussing open collaboration

### Practices of open collaboration

Across the papers found in our literature search, X out of 11 discuss actual experiences on implementing open collaboration, albeit in different organizational levels. These range from lab level (Turoman et al) center level (Bush et al and Grange et al), multi-center level (ManyPrimates et al and Grange et al) and field level: psychology (Alessandorni et al), Spinal Cord injury (Torres-Espin et al) and neuroimaging (Niso et al). All of these share in common the need for an infrastructure that promotes open science and open collaboration via starting building a community of collaborators or making tools (usually online resources) that researchers can use to reach FAIR (Finndable, Accessible, Interoperable and Reusable) goals while doing research. Before delving in those, here we provide the specific needs for implementing open collaboration in each organizational level, as well as describing the differences in approaches among experiences between and across levels (Figure 2).

#### Implementation of open collaboration practices at laboratory level

#### Implementation of open collaboration practices at center level

For center-level only one paper was identified (Bush et al), providing the example of a center of neuroimaging describing first how they identified the needs for open collaboration, then the individual-level practices that researchers should adapt and then the center-level changes to promote reaching FAIR goals.

#### Implementation of open collaboration practices at multi-center level

Interestingly, the level of organization with most papers is the multi-center level with two papers from ManyPrimates et al (ManyPrimates 2019 and ManyPrimates 2021) and Grange et al. In this organization level the most urgent needs are: 1) identifying the needs for open collaboration and 2) creating an infrastructure that allows those needs to be satisfied. Indeed, Grange et al describes a systematic review as an exploration of the needs for 11 UKRN centers to implement open collaborative practives. In this paper they specially focus on tools and resources among the 11 institutions and make clear the need for starting to utilize open resources as a first step to adapt open collaborative practices. The first paper from ManyPrimates et al [REF] describes the start of another multi-center collaboration, albeit from different institutions. It is interesting to observe that in this situation, the complex issue is to start the collaboration itself, which was ignited thanks to a symposium and an email chain among future collaborators. In contrast, ManyPrimates could quickly develop an open collaborative structure with relative ease in comparison to changing an already solid infrastructure, like in the case of Grange et al.

#### Implementation of open collaboration practices at field level

### Useful tools for open collaboration

## Discussions

## Conclusions

@@

We originally had another author involved, HC, who is on the protocol
author list. Since the time we uploaded the protocol, she has moved out
of academia and could no longer participate in any work.

Otherwise, the remaining authors (MG, DBI, and LWJ) had the following
contribution roles, following the [CRediT](https://credit.niso.org/)
taxonomy of conceptualization, data curation, formal analysis,
investigation, methodology, writing - original draft, writing -
reviewing and editing, project administration, software, validation, and
visualization.

>>>>

---
title: "Towards open collaboration in biomedical and health research: An open collaborative scoping review"
date: last-modified
---

@@

The full protocol for this scoping review was uploaded to the Open
Science Framework (DOI:
[`10.17605/OSF.IO/K9DR5`](https://doi.org/10.17605/OSF.IO/K9DR5)).

<!-- TODO: convert this to reference -->

### Formal sources

### Informal sources

<!-- TODO: need to add how we searched online sources, like lab websites -->

@@

### Deviations from the protocol

-   **Original timeline**: We were quite ambitious with the timeline for
    completing this scoping review. We did not adhere to the timeline.
-   **API vs web search results**: Almost all source databases had
    exposed APIs, with R packages available to access them. We chose to
    use these rather than using the web to extract the source data is
    due to our secondary aim. However, we noticed that there were some
    small differences in search results between the API query compared
    to the web-based query. We can't explain these differences, though
    they were quite minor.
-   **Exclude Web of Science source**: We were unable to connect the Web
    of Science source to R through the `{wosr}` package. Using the API
    key did not as documented in the package did not work. We decided
    that in the interest of reproducibility, that we exclude this
    source.
-   **Move all searching to using Open Alex**

@@

-   HC:
-   MG:
-   DBI: Conceptualization, Data curation, formal analysis,
    investigation, methodology, writing
-   LWJ:
